---
title: "Macroalgae condition on marine hard bottom"
subtitle: |
  [HARD_001] (lower growth limit, Skagerrak) \
  [HARD_002] (combined index, Nordsjøen–Barentshavet)
indicatorID:
  - HARD_001
  - HARD_002
indicatorName:
  - Macroalgae lower growth limit
  - Macroalgae combined littoral and sublittoral index
verbatimName:
  - Nedre voksegrenseindeks for makroalger (MSMDI)
  - Komboindeksen for makroalger
folderName: HARD_001-002 # Leave to be filled out by ecRxiv during typesetting (name of the folder under 'indicators/').
country: Norway
continent: Europe
ECT: B2 - Structural state characteristics
Realm:
  - Marine (M)
Biome:
  - M1 Marine shelf biome
Ecosystem:
  - M1.6 Subtidal rocky reefs; M1.2 Kelp forests
  - M1.6 Subtidal rocky reefs; M1.2 Kelp forests; MT1.1 Rocky shorelines
verbatimEcosystem: Marin hardbunn - fjæresone og sjøsone
yearAdded: 2026
yearLastUpdate: 2026
status: incomplete # options are complete, incomplete, or deprecated
Version: "0.1.0" # See wiki page: https://github.com/NINAnor/ecRxiv/wiki#versioning
VersionComment: First submission. Two indicators, one document.
url: https://github.com/NINAnor/ecRxiv/tree/main/indicators/HARD_001-002
Normalised: Yes
SpatialAggregationPathway: Scale - Aggregate (arithmetic mean)
Precision: Descriptive. Range and interquartile range across monitoring stations within a year. No inferential uncertainty is estimated, because each station-year yields a single published index value without replication.
format:
  html:
    embed-resources: true
    code-fold: true
    toc: true
    toc-title: Contents
    toc-depth: 3
    smooth-scroll: true
execute:
  cache: true
author:
  - name: Dag Ø. Hjermann
    email: dag.hjermann@niva.no
    affiliations:
      - id: 1
        name: Norwegian Institute for Water Research (NIVA)
AuthorList: Hjermann, D.Ø.
date: September 16, 2026
data_availability: none
code_reproducibility: none
open_science_badge: none
callout-icon: false
lightbox: true
css: ../../../style.css
code-links:
      - text: Add a review
        icon: github
        href: https://github.com/NINAnor/ecRxiv
hide: FALSE
bibliography: references.bib
codeReviewers:
  - name: To be assigned
    affiliations:
      - ref: 1
    reviewType: code review
---

<!--# This is a template for how to document the indicator analyses. Make sure also to not change the order, or modify, the headers, unless you really need to. This is because it easier to read if all the indicators are presented using the same layout. If there is one header where you don't have anything to write, just leave the header as is, and don't write anything below it. If you are providing code, be careful to annotate and comment on every step in the analysis. Before starting it is recommended to fill in as much as you can in the metadata file. This file will populate the initial table in your output.-->

<!--# Load all you dependencies here -->

```{r setup}
#| include: false
#| cache: false
library(knitr)
library(tidyverse)
library(kableExtra)
library(here)
library(anybadger)
library(yaml)
library(tibble)
library(conflicted)
# weird workarounds for cache issues
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
pull<-dplyr::pull

knitr::opts_chunk$set(echo = TRUE)
```

<!--# The following parts are autogenerated. Do not edit. -->


```{r source}
#| echo: false
#| cache: false
source(here::here("_common.R"))
```

```{r reviewers}
#| echo: false
#| results: asis

if (!is.null(yaml_data$codeReviewers)) {
  reviewers <- purrr::map_chr(yaml_data$codeReviewers, "name")

  htmltools::HTML(paste0(
    '<div id="title-block-header" class="quarto-title-block default">
       <div class="quarto-title-meta">
         <div>
           <div class="quarto-title-meta-heading">Reviewed by</div>
           <div class="quarto-title-meta-contents">',
    paste(reviewers, collapse = "; "),
    '</div>
         </div>
       </div>
     </div>',
    sep = ""
  ))
}
```

::: {layout-ncol="6"}

```{r statusBadge}
#| echo: false
status_badge(as.character(st))
```

```{r versionBadge}
#| echo: false
version_badge(my_version_number = as.character(version[[1]]))
```

```{r dataBadge}
#| echo: false
data_badge(as.character(badges[[1]]))
```

```{r codeBadge}
#| echo: false
code_badge(as.character(badges[[2]]))
```

```{r openScienceBadge}
#| echo: false
open_science_badge(as.character(badges[[3]]))
```

```{r licenseBadge}
#| echo: false
license_badge()
```


:::


> **Recommended citation**: `r paste(auth, " ", year, ". ", name, " (ID(s): ", folder_name, ") ", "v. ", version, ". ecRxiv: ", url, sep="")`



::: {.callout-tip collapse="true"}
## Metadata
```{r tbl-meta2}
#| tbl-cap: 'Indicator metadata'
#| echo: false
#| warning: false
meta |>
  dplyr::filter(Variable %in% c(
    "Indicator ID",
    "Indicator Name",
    "verbatimName",
    "Continent",
    "Country",
    "Ecosystem Condition Typology Class",
    "Realm",
    "Biome",
    "Ecosystem",
    "verbatimEcosystem",
    "Year added",
    "Last update",
    "Version",
    "Version comment",
    "Normalised",
    "Spatial aggregation pathway",
    "Precision")) |>
  kable()

```
:::

::: {.callout-tip collapse="true"}

## Log

<!--# Update this logg with short messages for each update -->
- 16 Sep. 2026 - Original PR. Open science badges left for ecRxiv to assign.
:::


<hr />

<!--# Document you work below.  -->

## 1. Summary

<!--#
With a maximum of 300 words, describe the indicator in general terms as you would to a non-expert. Think of this as a kind of common language summary. Follow the four numbers sub-headers below. under point 2, it is possible under  include a bullet point list of the specific steps in the workflow.
 -->

1. **Background and rationale**:
As sessile organisms, seaweeds and kelp are sensitive to changes in light conditions. How deep they grow is set by how much light reaches the seabed, and light penetration is reduced by phytoplankton, whose biomass is affected by nutrient runoff from waste-water treatment plants as well as nutrient runoff, especially form agricultural land. Light penetration is also affected by runoff of chemical constituents of terrestrial origin (humus). Seaweeds and kelp are also smothered by fast-growing opportunistic filamentous algae. A shallower lower growth limit, and a shift from perennial to opportunistic species, are therefore direct signals of eutrophication. 
2. **Methods**:
Two indicators are documented here, because Norway uses two different indices: 1. HARD_001 is based on the lower growth limit index (MSMDI), used in the Skagerrak, where the littoral (tidal) zone is very narrow). It scores the depth at which each of nine easily recognisable erect algae stops growing against reference depths which depend on water type. HARD_002 is based on the combination index (Komboindeksen), which is used along the rest of the coast, where the littoral zone is much more extensive). It averages a littoral species-composition index with a sublittoral score built from the depth limits of kelp and erect red algae, and the depth extent of filamentous algal mats. The indicators come from the national monitoring programmes ØkoKyst and Ytre Oslofjord which are part of Norway's obligations undet he Water Framework Directive (WFD), and both are expressed as an Ecological Quality Ratio (EQR) between 0 and 1. 
3. **Interpretation**:
An indicator value of 1 means the macroalgal community is minimally affected by human influence. Values decreasing toward zero mean algae are confined to progressively shallower water and opportunistic species take over. Values above 0.6 correspond to "good" or better ecological status under the Water Framework Directive.
4. **Results and development needs**:
Most stations are above 0.8 (thus in "good" condition in WFD terms), with a tendency toward lower values in 2024 than in 2021–2022. Both indicators are operational but incomplete. HARD_002 rests on only two of its three prescribed sublittoral sub-parameters, because the third is not reported to the national database. Neither indicator can yet be aggregated to water body for the whole station network. For HARD_001 the same station-year can carry two different official values depending on how far back the historical species record reaches; a revision of the calculation methodology is in preparation, and until it lands the longer time series should not be treated as method-consistent.

## 2. About the underlying data

<!--# Give a general introduction the datasets you have used, and provide more specific details in the sub-chapters.  -->

The data come from two national monitoring programmes: **ØkoKyst** (*Økosystemovervåking i kystvann*, activity code `OEKK`), run for the Norwegian Environment Agency, and **Ytre Oslofjord** (`YOFJ`), run for Fagrådet for Ytre Oslofjord. Both survey fixed hard-bottom stations along the coast and report to the national database **Vannmiljø**.  

Not everything the two indices need is in Vannmiljø, so three sources are combined:

| Source | What it provides |
|---|---|
| Vannmiljø registration export | raw species and depth registrations, plus the reported littoral index (RSLA/RSL) and, for Ytre Oslofjord, the reported MSMDI |
| ØkoKyst regional report tables for 2024 | the sublittoral sub-parameters of Komboindeksen, which are absent from Vannmiljø, and the MSMDI time series for the Skagerrak |
| ØkoKyst station overview | water body (*vannforekomst*) per station, which the Vannmiljø export does not carry |

### 2.1 Spatial and temporal resolution and extent

The observational unit is the monitoring station — a stretch of shore of 8–15 m for the littoral survey, and a dive or camera transect running from the shoreline down to about 30 m for the sublittoral survey. Stations are revisited, so each station contributes one index value per survey year.

We have focused on 2024, as that is the last year covered in published reports. We have looked at Vannmiljø data exported for 2021–2024, chosen to match the start of the current programme period (2021-2025). HARD_001 covers 33 Skagerrak stations over four years, 83 station-years in all. HARD_002 covers 18 surveyed stations from Nordsjøen to Barentshavet covers only for 2024 (sublittoral sub-parameters were obtained for only that year).

The Skagerrak report tabulates MSMDI back to 2009, but the series is deliberately cut at 2021 rather than extended. @lundsor2025 document that the 2021–2024 values are computed by the current method, while earlier years come from a mix of sources and two different calculation conventions — some from individual annual reports, some from manual registrations in Vannmiljø that probably used a three-year lookback, one value corrected for a reporting error, and the 2009–2016 values from a later recalculation. Splicing those together would produce an apparent trend that partly reflects changes in method rather than in condition.

### 2.2 Sampling design

Stations are placed on continuous hard bottom (bedrock or stable boulder), sloping evenly to 30 m or more, and areas with conspicuous ice or sand scouring are avoided. This is a targeted design where stations are deliberately chosen to be representative of a water body and to be comparable between years, not drawn at random. 

The littoral survey for HARD_002 is done by snorkelling between mid and high tide, recording all attached macroalgae on a six-point cover scale. The sublittoral survey is done by diving for MSMDI — a drop camera is explicitly not accepted, because several of the nine index species cannot be identified reliably from video — and by ROV or drop camera for Komboindeksen, with 3–5 replicate transects per station.

### 2.3 Original units

The measurement of parameters and the calculation of indices are described in *Veileder 02:2018* [@veileder022018], the monitoring guidelines for Norwegian monitoring under the EU Water Framework Directive. The MSMDI index (HARD_001) is a formal classification tool described in the monitoring guidelines, and so is the littoral index (fjæreindeksen), which is half of the komboindeks. The komboindels itself is not yet a formal classification tool. 

| Variable | Original unit |
|---|---|
| Lower growth limit (`MANVGR`) | metres below chart datum, one value per taxon |
| Relative abundance along transect (`MARFORSK`) | ordinal 1–4 (*enkeltfunn*, *spredt*, *vanlig*, *dominerende*), recorded per 1 m depth interval |
| Littoral cover (`MADEKNSK`) | ordinal 1–6 cover classes |
| Depth extent of filamentous mats | metres (summed over depth intervals with >50 % cover) |
| Reported indices (`MSMDI1-3`, `RSLA1-3`, `RSL4-5`) | EQR, dimensionless, 0–1 |

### 2.4 Instructions for citing, using and accessing data

The Vannmiljø data can be downloaded from the Norwegian Environment Agency's database at <https://vannmiljo.miljodirektoratet.no/> and are licensed under Norsk lisens for offentlige data (NLOD) 2.0, <https://data.norge.no/nlod>. The filtered extract used here is included in `data/`, with provenance in `data/README_data.md`.

The values taken from the ØkoKyst 2024 regional reports are not in a machine-readable public archive. They were transcribed from the report tables into `data/kombo_reported_okokyst.csv` and `data/msmdi_reported_okokyst.csv`; each row records its source table. The transcription was checked arithmetically and, where the same value also exists in Vannmiljø, against that value (§5.1).

### 2.5 Additional comments about the dataset

- Index values are often duplicated. Each station-year often carries the index twice, once rounded with `Provetakmetode_id = "NS-EN ISO 19493:2007 SO"` (or `... NV` for lower growth limits) and once unrounded with method `UKJENT`. The two sometimes differ by more than rounding. We use the rounded record, falling back to the unrounded one rounded to two decimals where no rounded record exists.
- `Vannlokalitet_kode`, not the station name, is the key, as the station name can vary slightly in spellin. Also, the same station name can occur under two different codes, and the short station code `Betegnelse` is missing for most Ytre Oslofjord records.

## 3. Indicator properties

### 3.1 Ecosystem Condition Typology Class (ECT)

Both indicators are assigned to **B2 – Structural state characteristics**. What they measure is the vertical extent of the macroalgal community: how far down the seabed the perennial algae reach. That is a structural property of the habitat rather than a chemical or functional one.

The assignment is not perfectly clean for HARD_002. Half of Komboindeksen is the littoral index RSLA/RSL, which scores species composition — proportions of green, red and brown algae, and the ratio of perennial to opportunistic species — and that half would sit more naturally in B1 – Compositional state characteristics. We classify the combined indicator by its sublittoral half and by the pressure it is designed to detect.

### 3.1.1 Other condition typologies

Both indices are formal classification tools under the EU Water Framework Directive, implemented in Norway through *Veileder 02:2018* [@veileder022018]. There, they are biological quality elements for the pressure type eutrophication, and the EQR scale maps onto the five WFD status classes. The class boundaries used here are the WFD ones, so indicator values are directly comparable with official water body status assessments.

### 3.2 Ecosystem condition characteristic

Depth extent of perennial macroalgae.

### 3.3 Collinearities with other indicators

Both indicators respond to light attenuation, so they are expected to correlate with any indicator built on Secchi depth, chlorophyll *a*, or total nitrogen and phosphorus in coastal water. Including one of those alongside these indicators in the same assessment would double-count the eutrophication signal.

There is also a partial overlap between the two indicators documented here. HARD_002's sublittoral half uses the lower growth limit of erect red algae and kelp, which is the same class of measurement as HARD_001. They are not collinear in practice only because they apply to different ecoregions and are never computed for the same station.

A weaker overlap exists with eelgrass indicators, which respond to the same pressure through the same mechanism but on soft bottom.

### 3.4 Impact factors

The pressure the indices are designed for is eutrophication: nutrient inputs from agriculture, aquaculture, and wastewater reduce light by increasing plankton biomass and by promoting filamentous algal mats that overgrow perennial species.

Several other factors change the metric without being eutrophication, and this is the main interpretive difficulty:

- **Sea urchin grazing** can remove the macroalgal community outright. Because grazing suppresses the very filamentous-algae response the index reads as a eutrophication signal, a heavily grazed station can score deceptively well on parts of the index while being in poor condition. A correction for this was proposed by @fagerli2023 and is included in the revised method, but we could not implement it (§5.1).
- **Natural turbidity** from river runoff and resuspension reduces light without any anthropogenic cause. Water type partly accounts for this, since reference depths differ between exposed coast and freshwater-influenced fjord.
- **Substrate**: where hard bottom gives way to sediment above the light-limited depth, the recorded limit reflects the substrate, not the light. The method excludes such observations.
- **Climate-driven loss of sugar kelp** in southern Norway, driven by warming and reduced water exchange, affects one of the nine MSMDI species.

### 3.5 Spatial aggregation

The reporting unit is the monitoring station in a given year, and no spatial averaging is applied to produce it. Within a station-year the aggregation pathway is *Scale – Aggregate (arithmetic mean)*:

- the observed variable (a depth in metres, or a depth extent in metres) is **scaled** against water-type-specific reference levels to a points score, and divided by the reference score of 5 to give a sub-parameter EQR;
- sub-parameter EQRs are **aggregated** by unweighted arithmetic mean — over the nine species for HARD_001, and over three sublittoral sub-parameters and then the two index halves for HARD_002.

No transformation and no truncation are applied; the point scale bounds the result to 0–1 by construction.

*Veileder 02:2018* defines a further step we do **not** take: the normalised EQR of a **water body** is the mean of its station EQRs. We stop at the station because the Vannmiljø export carries no water body identifier. `data/station_waterbody.csv` supplies one for the ØkoKyst stations, but not for the 17 Ytre Oslofjord stations, so a complete water-body series is not yet possible (§5.1).

## 4. Reference
### 4.1 Reference condition

The reference condition is the macroalgal community expected at a station of its water type under **minimal human influence**: perennial algae extending to the depth that natural light attenuation allows, a species composition dominated by late-successional perennials, and no mass occurrence of filamentous algae.

It is defined per **ecoregion and water type** rather than globally, because natural light climate, salinity and wave exposure differ enormously between an exposed Skagerrak coast and a freshwater-influenced fjord in Troms. Reference depths were derived from historical records, data from unimpacted areas, and expert judgement [@gundersen2017].

### 4.2 Reference levels

Both indices are normalised on a **points scale anchored at both ends**. X~100~ is the reference depth for the taxon and water type, which earns the maximum score; X~0~ is absence of the taxon (or, for filamentous algae, the largest depth extent class), which earns zero. The EQR is the mean score divided by the maximum score of 5, so the indicator is bounded 0–1 with 1 as reference condition.

The two indices use **different point scales**, which matters when comparing them:

- **HARD_001 (MSMDI)** scores 5, 4, 3, 2 or 0 points per species, from four depth thresholds per species and water type. Zero is given only if the species was recorded at the station previously but not in the survey year.
- **HARD_002 (Komboindeksen)** scores 5, 4.5, 3.5, 2.5, 1.5 or 0 points per sub-parameter, from five thresholds.

X~60~, the good/moderate boundary, is 0.6 on the EQR scale for both, which is the WFD boundary. The five status classes are equal 0.2-wide bands.

The full threshold tables are reproduced in the analysis code in §9, with their sources: *Veileder 02:2018* Tables 9.4–9.6 for HARD_001, and Fagerli et al. (2025) Tables 3–6 for HARD_002 [@fagerli2025].

#### 4.2.1 Spatial resolution and validity

Reference levels vary by **ecoregion and water type**, and are therefore valid for scaling a value measured at a single station, provided the station's water type is known. This is convenient — no prior spatial aggregation is needed before scaling — but it makes the water type a critical piece of metadata, and it is not carried in the Vannmiljø registration export. We recover it from the reported index parameter code, which encodes it (`MSMDI3` means water type 3), and confirm it against the ØkoKyst Skagerrak report.

Coverage of the reference levels is incomplete, and the gaps are filled by substitution in the source reports rather than by us:

- HARD_001 thresholds exist for Skagerrak water types 1–3 only, which is why the indicator is restricted to that ecoregion.
- HARD_002 kelp thresholds do not exist for Barentshavet, and for Norskehavet only for water type 1; the reports substitute values from a neighbouring water type or ecoregion. Every affected station is flagged in the data.

## 5. Uncertainties
### 5.1 General uncertainties

The largest uncertainties here are not statistical. They concern which published number to trust and how much of the prescribed method could actually be executed.

**Two official MSMDI series disagree, and the reason is documented.** For the ØkoKyst Skagerrak stations, the values held in Vannmiljø and the values published in the ØkoKyst Skagerrak 2024 report differ systematically: the report is on average 0.12 lower, up to 0.35, and **27 of 42 overlapping station-years fall in a different condition class** depending on which is used.

@lundsor2025 address this directly. The cause is the rule that gives 0 points to a species previously recorded at a station but not refound in the survey year. How far back "previously" reaches depends on what data exist:

- Vannmiljø computes MSMDI automatically from reported lower growth depths and **cannot look further back than the earliest registrations in the system** — about 2009 for most stations, but not all, and in some years only EQR values were reported rather than the underlying depths.
- NIVA changed practice in report M-1336|2019 to look only **three years back**, which the report notes yields potentially higher MSMDI values and a better status class than the *Veileder 02:2018* method would.

The report states that its **2021–2024 values are computed by the current method using previous occurrences registered in Vannmiljø, i.e. back to 2009 for most stations**. That is the longest lookback available, so we use the report series. It is consistent with our own reimplementation, which can only look back to 2021 and accordingly sits slightly *above* the report (mean bias +0.02) while sitting well below Vannmiljø (−0.10); fewer years of lookback means fewer zeros and a higher score. §9.4 quantifies this.

This is the single largest methodological choice in HARD_001. @lundsor2025 also note that a revision of the calculation methodology is in preparation, and recommend that an agreed reference list of historical species occurrences be established and made available to Vannmiljø's automatic calculation. Until that happens, MSMDI values from different sources and periods are not strictly comparable.

**HARD_002 never uses all three sublittoral sub-parameters.** The depth extent of filamentous algal mats is not reported to Vannmiljø for any station in any year, so every value rests on the two depth-limit sub-parameters plus a filamentous-algae score taken from the report tables. The method permits averaging over the sub-parameters that are measurable, stating explicitly that statistical power is reduced.

**HARD_002 mixes two point scales.** Nordsjøen and Norskehavet sør use the revised half-point scale of @fagerli2025; Norskehavet nord and Barentshavet use an older integer scale. Five of the 13 stations with a value are on the older scale — 10 of the 18 stations surveyed. Values are not strictly comparable between these regions, and each row in the data records which scale applies.

**The sea urchin grazing correction is not implemented, and cannot be approximated well.** The method downgrades a station by one class where urchins are at least "spredt" over more than 5 m of grazed depth. Vannmiljø has urchin abundance but no parameter for grazed depth extent, so the correction cannot be computed. We flag urchin abundance in §9.5 instead, but the flag is weak evidence: **77 of the 83 HARD_001 station-years and 10 of the 13 HARD_002 stations have no urchin record at all**, because the relevant parameter is only registered at a subset of stations. Absence of a flag therefore does not mean absence of grazing.

**The "disappeared species" rule cannot be applied fully in our reimplementation.** We can only see absences within the 2021–2024 extract, whereas the published values use occurrences back to about 2009. Species lost earlier are invisible to us and are simply excluded rather than scored zero, which biases our reimplementation upward by roughly the +0.02 seen in §9.4. This affects the quality check, not the published values. It is the same mechanism that @lundsor2025 identify as the source of the discrepancy between series, so it is a property of the index as operated nationally rather than a defect of this workflow.

**Targeted station placement** (§2.2) means values describe the surveyed localities, not hard bottom at large.

**No water-body aggregation** is possible for the Ytre Oslofjord stations, so the station-level series cannot yet be rolled up consistently.

### 5.2 Reported uncertainties

What we report alongside the point estimates in @tbl-results is **descriptive, not inferential**: the range and interquartile range of station values within a year. It describes spatial variation between monitoring stations, not the uncertainty of any single station value.

Each station-year yields exactly one published index value, with no replication at the station level, so there is no basis for an inferential interval. The sub-parameters that go into it are ordinal scores, and the index is their unweighted mean; a standard error of that mean would be a statement about between-species or between-sub-parameter variation, which is not a measurement error and would be misleading to present as one.

The agreement between our reimplementation and the published values, reported in §9.4, is the closest thing to an accuracy estimate available: for the ØkoKyst half of HARD_001, 29 of 48 station-years agree within 0.05 EQR units, with a mean bias of +0.02.

## 6. References

<!--# You can add references manually or use a citation manager and add intext citations as with crossreferencing and hyperlinks. See https://quarto.org/docs/authoring/footnotes-and-citations.html -->

::: {#refs}
:::

## 7. Import and prepare datasets

<!--# All the data import and preparation should be done in this chapter, before the analysis chapter. As a minimum, add code for importing data, and describe or show the data structure. You may also explain certain variables, what they represent, and their distributions -->

All files are UTF-8 with `.` as decimal separator. The encoding is stated explicitly on read, so the result does not depend on the reader's locale — this matters because the data contain Norwegian station names and the files were prepared on a Norwegian-locale Windows machine.

```{r read-helper}
# All csv files live in data/, one level up from this qmd.
read_data <- function(file) {
  readr::read_csv(
    file.path("..", "data", file),
    locale = readr::locale(encoding = "UTF-8"),
    show_col_types = FALSE, progress = FALSE
  )
}
```

### 7.1 Vannmiljø registrations

The macroalgae registrations, filtered from the full Vannmiljø export to the parameters the two indices need. One row is one registration: a value for one parameter, at one station, on one date, optionally for one taxon at one depth.

```{r read-vannmiljo}
macroalgae <- read_data("vm_macroalgae.csv")
stations   <- read_data("vm_stations.csv")
parameters <- read_data("vm_parameters.csv")

# Station key is Vannlokalitet_kode throughout; see section 2.5
key <- c("Vannlokalitet_kode", "year")

tibble(
  rows       = nrow(macroalgae),
  stations   = n_distinct(macroalgae$Vannlokalitet_kode),
  parameters = n_distinct(macroalgae$Parameter_id),
  years      = paste(range(macroalgae$year), collapse = "-")
) |> kable(caption = "Vannmiljø macroalgae extract")
```

The parameters present, and what each contributes:

```{r tbl-parameters}
macroalgae |>
  count(Parameter_id, name = "records") |>
  left_join(parameters, by = "Parameter_id") |>
  filter(!str_detect(Parameter_id, "^(DF|AF|BF)")) |>
  arrange(desc(records)) |>
  select(Parameter_id, Parameter_navn, Enhet, records) |>
  kable(caption = "Macroalgae parameters in the extract. The DF*/AF*/BF* habitat description parameters are omitted from this table for brevity.")
```

### 7.2 Published index values not held in Vannmiljø

Two tables transcribed from the ØkoKyst 2024 regional reports. Without these, HARD_002 cannot be computed at all, and HARD_001 would be missing 2023 entirely.

```{r read-reports}
kombo_reported <- read_data("kombo_reported_okokyst.csv")
msmdi_reported <- read_data("msmdi_reported_okokyst.csv")

# HARD_002 sub-parameters: one row per station-year. Rows for years before the
# 2024 survey carry only the final index, so they are not usable here.
kombo_reported |>
  filter(year == 2024) |>
  count(ecoregion, point_scale, name = "stations") |>
  kable(caption = "Komboindeksen sub-parameters from the 2024 reports, by ecoregion and point scale")
```

```{r tbl-msmdi-source}
msmdi_reported |>
  count(year, in_vannmiljo, name = "station_years") |>
  pivot_wider(names_from = in_vannmiljo, values_from = station_years,
              values_fill = 0, names_prefix = "in_vannmiljo_") |>
  kable(caption = "MSMDI station-years from the Skagerrak report, and whether the same station-year also exists in Vannmiljø")
```

### 7.3 Station metadata

Water body and ecoregion per station, from the ØkoKyst station overview for 2026–2030. Note that the workbook's own `Vanntype` column is a **Vann-Nett** water type code, which is a different typology from the 1–6 water types the index class boundaries use — it agreed with the index water type for only 2 of 91 stations. We therefore use this file for the water body and the ecoregion only.

```{r read-stationmeta}
station_meta <- read_data("station_waterbody.csv")

station_meta |>
  summarise(
    stations     = n(),
    water_bodies = n_distinct(water_body, na.rm = TRUE),
    with_ecoregion = sum(!is.na(ecoregion))
  ) |> kable(caption = "Station metadata coverage (ØkoKyst stations only)")
```

## 8. Spatial units

The **Basic Spatial Unit** is the monitoring station: a shoreline stretch of 8–15 m together with the transect running from it down to about 30 m. Indicator values are valid, and should be interpreted, at this unit.

The **Ecosystem Asset** that the station is taken to represent is the *vannforekomst* (water body), the management unit of the Water Framework Directive, and *Veileder 02:2018* defines water body status as the mean of the station values within it. `data/station_waterbody.csv` provides this mapping for the ØkoKyst stations, and is used in §10 to show what a water-body series would look like. It is not available for the Ytre Oslofjord stations, so we report at station level and treat water-body aggregation as a development need.

The **Ecosystem Accounting Area** would be the ecoregion (Skagerrak, Nordsjøen sør/nord, Norskehavet sør/nord, Barentshavet), which is also the unit the reference levels are defined for. No masking layer of hard-bottom extent is used: the indicators describe the surveyed stations, and scaling them to area would require a coastal substrate map that is outside the scope of this submission.

## 9. Analyses

<!--#
Use this header for documenting the analyses. Put code in separate code chunks, and annotate the code in between using normal text (i.e. between the chunks, and try to avoid too many hashed out comments inside the code chunks). Add subheaders as needed.
-->

### 9.1 Shared definitions

The five status classes are equal 0.2-wide bands on the EQR scale, and are identical for both indices — *Veileder 02:2018* Table 9.7 for MSMDI, Fagerli et al. (2025) Table 2 for Komboindeksen.

```{r class-function}
# Condition class from an EQR value. Labels are the Norwegian WFD classes.
neqr_class <- function(x) {
  cut(x, breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1),
      labels = c("Svært dårlig", "Dårlig", "Moderat", "God", "Svært god"),
      include.lowest = TRUE, right = TRUE)
}

# Colours prescribed by the Norwegian classification system. These are a status
# palette: they encode state, not series identity, and are always accompanied by
# the class label so that the information is never carried by colour alone.
class_colours <- c(
  "Svært dårlig" = "#E4181C",
  "Dårlig"       = "#F59331",
  "Moderat"      = "#FDE64B",
  "God"          = "#5DBB63",
  "Svært god"    = "#1E90D2"
)
```

The littoral index RSLA/RSL is taken as reported rather than recomputed. It is a six-parameter multimetric index requiring species lists with ecological status group and opportunist attributes, plus a habitat-based normalisation factor; reimplementing it was judged out of scope, and it is reported to Vannmiljø for every relevant station. The extraction below applies the de-duplication rule from §2.5.

```{r fjaeresone}
# Fjæresoneindeks (RSLA1-3, RSL4-5) per station-year, de-duplicated.
# A station can be surveyed twice in a year (one is, in 2022), in which case the
# surveys are averaged; and a few station-years are reported under two water
# types, in which case the water type is left NA and flagged.
fjaeresone <- macroalgae |>
  filter(str_detect(Parameter_id, "^RSL")) |>
  mutate(variant = if_else(Provetakmetode_id == "NS-EN ISO 19493:2007 SO",
                           "rounded", "unrounded")) |>
  group_by(across(all_of(key)), rsl_param = Parameter_id, variant) |>
  summarise(value = mean(value), n_surveys = n_distinct(date), .groups = "drop") |>
  pivot_wider(names_from = variant, values_from = c(value, n_surveys),
              names_glue = "{.value}_{variant}") |>
  mutate(eqr_fjaeresone = coalesce(value_rounded, round(value_unrounded, 2)),
         water_type     = as.integer(str_remove(rsl_param, "^RSLA?"))) |>
  group_by(across(all_of(key))) |>
  summarise(
    rsl_param      = paste(sort(unique(rsl_param)), collapse = "/"),
    water_type     = if (n() == 1) water_type[1] else NA_integer_,
    eqr_fjaeresone = mean(eqr_fjaeresone),
    n_surveys      = max(c(n_surveys_rounded, n_surveys_unrounded), na.rm = TRUE),
    .groups = "drop"
  )

glimpse(fjaeresone)
```

### 9.2 HARD_001 — lower growth limit index (MSMDI)

MSMDI is the mean score of nine easily recognisable erect algae, scored by the depth at which each stops growing. The nine species occupy **eight scoring rows**, because krusblekke (*Phyllophora pseudoceranoides*) and hummerblekke (*Coccotylus truncatus*) share a row and the deeper of the two is used.

The indicator value is a **published** index value, not our own computation, and the source is chosen per programme for the reason given in §5.1:

```{r hard001}
# ØkoKyst: the Skagerrak 2024 report series. Lundsør et al. (2025) state that
# its 2021-2024 values use the current method with previous species occurrences
# taken back to about 2009 - the longest lookback available, and longer than
# either Vannmiljø's automatic calculation or our reimplementation in 9.4.
hard001_oekk <- msmdi_reported |>
  transmute(Vannlokalitet_kode, year, station_code, Vannlokalitetsnavn,
            programme = "OEKK", water_type,
            eqr = msmdi_report,
            value_source = "ØkoKyst Skagerrak 2024 report, Tabell 5",
            eqr_vannmiljo = msmdi_vannmiljo)

# Ytre Oslofjord: Vannmiljø, since the Skagerrak report does not cover the
# Fagrådet network. Rounded record preferred, as in section 2.5. Water type is
# read off the parameter code, and preferring the rounded record settles the one
# station-year reported under two water types.
hard001_yofj <- macroalgae |>
  filter(str_detect(Parameter_id, "^MSMDI[123]$"), !is.na(value),
         Aktivitet_id == "YOFJ") |>
  mutate(variant    = if_else(str_detect(Provetakmetode_id, "NS-EN ISO 19493"),
                              "rounded", "unrounded"),
         water_type = as.integer(str_remove(Parameter_id, "^MSMDI"))) |>
  group_by(across(all_of(key)), variant) |>
  summarise(value = mean(value), water_type = first(water_type),
            .groups = "drop") |>
  pivot_wider(names_from = variant, values_from = c(value, water_type),
              names_glue = "{.value}_{variant}") |>
  mutate(eqr        = coalesce(value_rounded, round(value_unrounded, 2)),
         water_type = coalesce(water_type_rounded, water_type_unrounded),
         programme  = "YOFJ",
         value_source = if_else(!is.na(value_rounded),
                                "Vannmiljø, rounded record",
                                "Vannmiljø, unrounded record rounded to 2 dp"),
         eqr_vannmiljo = eqr) |>
  left_join(macroalgae |> distinct(Vannlokalitet_kode, Betegnelse) |>
              filter(!is.na(Betegnelse)) |> rename(station_code = Betegnelse),
            by = "Vannlokalitet_kode") |>
  left_join(stations |> select(Vannlokalitet_kode, Vannlokalitetsnavn),
            by = "Vannlokalitet_kode") |>
  select(all_of(key), station_code, Vannlokalitetsnavn, programme, water_type,
         eqr, value_source, eqr_vannmiljo)

hard001 <- bind_rows(hard001_oekk, hard001_yofj) |>
  mutate(indicatorID = "HARD_001", index = "MSMDI", ecoregion = "Skagerrak",
         condition_class = neqr_class(eqr)) |>
  arrange(programme, station_code, Vannlokalitet_kode, year)

# Guards: one value per station-year, on scale, water type within the range the
# index is defined for, and no station in both programmes.
stopifnot(
  !any(duplicated(paste(hard001$Vannlokalitet_kode, hard001$year))),
  all(!is.na(hard001$eqr)), all(hard001$eqr >= 0 & hard001$eqr <= 1),
  all(hard001$water_type %in% 1:3),
  max(count(distinct(hard001, Vannlokalitet_kode, programme),
            Vannlokalitet_kode)$n) == 1
)

hard001 |> count(programme, year, name = "stations") |>
  pivot_wider(names_from = year, values_from = stations, values_fill = 0) |>
  kable(caption = "HARD_001: stations per programme and year")
```

### 9.3 HARD_002 — combined index (Komboindeksen)

Komboindeksen averages the littoral index with a sublittoral score:

$$\text{EQR}_{\text{sjøsone}} = \frac{\text{EQR}_{\text{kelp}} + \text{EQR}_{\text{erect red algae}} + \text{EQR}_{\text{filamentous algae}}}{3}$$

$$\text{nEQR}_{\text{Kombo}} = \frac{\text{EQR}_{\text{fjæresone}} + \text{EQR}_{\text{sjøsone}}}{2}$$

Each sublittoral sub-parameter is scored 5, 4.5, 3.5, 2.5, 1.5 or 0 points from depth thresholds specific to ecoregion and water type, then divided by the reference score of 5. The kelp sub-parameter uses *Laminaria hyperborea* in exposed water type 1 and the deeper of *L. hyperborea* and *Saccharina latissima* in water types 2–4; the filamentous-algae sub-parameter is inverse, so zero depth extent is best.

We recompute the index rather than use the published value, because the four regional reports are **not mutually consistent**: they differ in which littoral value they feed in and in rounding. Recomputing applies one method to all four ecoregions. The sub-parameter EQRs themselves are taken as scored in the reports, since the underlying depths in metres are not available.

```{r hard002}
# Aggregation functions, exactly as in Fagerli et al. (2025) step 6-7. Both drop
# sub-parameters that were not measurable and average the rest, which the method
# permits with the caveat of reduced statistical power.
eqr_sjosone_fun <- function(kelp, red, filamentous) {
  e <- c(kelp, red, filamentous)
  if (all(is.na(e))) NA_real_ else mean(e, na.rm = TRUE)
}
neqr_kombo_fun <- function(fjaeresone, sjosone) {
  if (is.na(fjaeresone) || is.na(sjosone)) NA_real_ else mean(c(fjaeresone, sjosone))
}

hard002 <- kombo_reported |>
  filter(year == 2024) |>
  select(Vannlokalitet_kode, station_code, Vannlokalitetsnavn, ecoregion, year,
         eqr_tare, eqr_rodalger, eqr_tradalger, point_scale,
         unapproved_boundaries, neqr_kombo_published = neqr_kombo) |>
  left_join(fjaeresone |> select(all_of(key), rsl_param, water_type,
                                 eqr_fjaeresone), by = key) |>
  rowwise() |>
  mutate(
    n_subparameters = sum(!is.na(c(eqr_tare, eqr_rodalger, eqr_tradalger))),
    eqr_sjosone     = eqr_sjosone_fun(eqr_tare, eqr_rodalger, eqr_tradalger),
    eqr             = neqr_kombo_fun(eqr_fjaeresone, eqr_sjosone)
  ) |>
  ungroup() |>
  mutate(
    indicatorID = "HARD_002", index = "Komboindeksen", programme = "OEKK",
    condition_class = neqr_class(eqr),
    status = case_when(
      n_subparameters == 0    ~ "no sublittoral sub-parameter observed",
      is.na(eqr_fjaeresone)   ~ "no littoral index in Vannmiljø",
      TRUE                    ~ "computed"
    ),
    # Flags for section 5.1
    flag_old_point_scale = str_detect(point_scale, "^integer"),
    flag_partial_sjosone = n_subparameters < 3
  ) |>
  arrange(ecoregion, station_code)

stopifnot(
  !any(duplicated(paste(hard002$Vannlokalitet_kode, hard002$year))),
  all(hard002$eqr >= 0 & hard002$eqr <= 1, na.rm = TRUE),
  # Every computed value must rest on at least one sub-parameter
  all(hard002$n_subparameters > 0 | hard002$status != "computed")
)

hard002 |>
  count(ecoregion, status, name = "stations") |>
  kable(caption = "HARD_002: stations per ecoregion and computation status")
```

Because all three sublittoral sub-parameters were needed but the filamentous-algae one is nowhere in Vannmiljø, it is worth stating plainly how many stations rest on how much evidence:

```{r hard002-evidence}
hard002 |>
  filter(status == "computed") |>
  count(n_subparameters, flag_old_point_scale, name = "stations") |>
  rename(`sublittoral sub-parameters used` = n_subparameters,
         `older integer point scale` = flag_old_point_scale) |>
  kable(caption = "HARD_002: evidence base per station")
```

### 9.4 Quality check — independent reimplementation of MSMDI

This section is a check, not a source of published values. It reimplements MSMDI from the raw registrations and compares the result with the two official series, which is what justifies the choice made in §9.2.

The scoring table is transcribed from *Veileder 02:2018* Tables 9.4–9.6 and was verified against rendered images of those pages. Columns `p5`–`p2` are depth thresholds in metres: a species scores 5 points if its lower growth limit is deeper than `p5`, 4 if deeper than `p4`, and so on.

```{r msmdi-thresholds}
msmdi_thresholds <- tribble(
  ~taxon_group,             ~water_type, ~p5, ~p4, ~p3, ~p2,
  # MSMDI 1 - S1, open exposed coast
  "Chondrus crispus",                 1,  13,   9,   5,   0,
  "Furcellaria lumbricalis",          1,  12,   9,   5,   0,
  "Halidrys siliquosa",               1,  10,   8,   4,   0,
  "Saccharina latissima",             1,  12,   9,   5,   0,
  "Phyllophora/Coccotylus",           1,  22,  18,   9,   0,
  "Rhodomela confervoides",           1,  12,   9,   5,   0,
  "Delesseria sanguinea",             1,  22,  18,   9,   0,
  "Phycodrys rubens",                 1,  22,  17,   9,   0,
  # MSMDI 2 - S2, moderately exposed coast/fjord
  "Chondrus crispus",                 2,   8,   5,   3,   0,
  "Furcellaria lumbricalis",          2,  10,   7,   4,   0,
  "Halidrys siliquosa",               2,   8,   5,   3,   0,
  "Saccharina latissima",             2,  10,   7,   4,   0,
  "Phyllophora/Coccotylus",           2,  18,  12,   6,   0,
  "Rhodomela confervoides",           2,  12,   7,   4,   0,
  "Delesseria sanguinea",             2,  18,  12,   6,   0,
  "Phycodrys rubens",                 2,  15,  10,   5,   0,
  # MSMDI 3 - S3, sheltered coast/fjord
  "Chondrus crispus",                 3,  10,   7,   4,   0,
  "Furcellaria lumbricalis",          3,  12,   8,   4,   0,
  "Halidrys siliquosa",               3,  10,   7,   4,   0,
  "Saccharina latissima",             3,   8,   6,   3,   0,
  "Phyllophora/Coccotylus",           3,  10,   8,   4,   0,
  "Rhodomela confervoides",           3,  12,   8,   4,   0,
  "Delesseria sanguinea",             3,  13,   9,   5,   0,
  "Phycodrys rubens",                 3,  13,   8,   4,   0
)

# Vannmiljø scientific names to the eight scoring rows. Both spellings of
# Phyllophora pseudoceranoides occur, one with a diaeresis.
msmdi_taxon_map <- c(
  "Chondrus crispus"             = "Chondrus crispus",
  "Furcellaria lumbricalis"      = "Furcellaria lumbricalis",
  "Halidrys siliquosa"           = "Halidrys siliquosa",
  "Saccharina latissima"         = "Saccharina latissima",
  "Phyllophora pseudoceranoides" = "Phyllophora/Coccotylus",
  "Phyllophora pseudoceranoïdes" = "Phyllophora/Coccotylus",
  "Coccotylus truncatus"         = "Phyllophora/Coccotylus",
  "Rhodomela confervoides"       = "Rhodomela confervoides",
  "Delesseria sanguinea"         = "Delesseria sanguinea",
  "Phycodrys rubens"             = "Phycodrys rubens"
)
```

Lower growth limits reach Vannmiljø by two different routes, because the two programmes report differently. Ytre Oslofjord reports the limit directly; ØkoKyst reports only the transect abundance records, from which the limit is the deepest metre at which the taxon occurs at least *spredt* — the definition the veileder uses.

```{r msmdi-limits}
# Route A: lower growth limit given directly, one record per taxon.
route_a <- macroalgae |>
  filter(Parameter_id == "MANVGR", !is.na(value)) |>
  mutate(taxon_group = unname(msmdi_taxon_map[VitenskapligNavn])) |>
  filter(!is.na(taxon_group)) |>
  group_by(across(all_of(key)), taxon_group) |>
  summarise(depth_m = max(value), .groups = "drop") |>
  mutate(route = "A: direct")

# Route B: derived from the dive transect. Depths of -1 mark the littoral zone.
marforsk <- macroalgae |>
  filter(Parameter_id == "MARFORSK", !is.na(value), !is.na(upper_m))

route_b <- marforsk |>
  filter(value >= 2, upper_m >= 0) |>
  mutate(taxon_group = unname(msmdi_taxon_map[VitenskapligNavn])) |>
  filter(!is.na(taxon_group)) |>
  group_by(across(all_of(key)), taxon_group) |>
  summarise(depth_m = max(upper_m), .groups = "drop") |>
  mutate(route = "B: derived from transect")

growth_limits <- bind_rows(route_a, route_b)
```

Scoring then follows the veileder: points per species from the thresholds, zero for a species recorded at the station in an earlier year but not in the survey year, exclusion of species never recorded, and a minimum of three species for the index to be computed.

```{r msmdi-score}
# Species recorded at the station in an earlier year but not in this one score
# zero and still count in the mean. We can only see this within 2021-2024, which
# is why the reimplementation is biased slightly upward (section 5.1).
first_seen <- growth_limits |>
  group_by(Vannlokalitet_kode, taxon_group) |>
  summarise(first_year = min(year), .groups = "drop")

disappeared <- growth_limits |>
  distinct(Vannlokalitet_kode, year) |>
  inner_join(first_seen, by = "Vannlokalitet_kode",
             relationship = "many-to-many") |>
  filter(year > first_year) |>
  anti_join(growth_limits, by = c("Vannlokalitet_kode", "year", "taxon_group")) |>
  transmute(Vannlokalitet_kode, year, taxon_group, depth_m = NA_real_,
            disappeared = TRUE, route = "disappeared")

msmdi_obs <- bind_rows(growth_limits |> mutate(disappeared = FALSE),
                       disappeared) |>
  # Water type from the published series, the only place it is recorded
  left_join(bind_rows(
    msmdi_reported |> distinct(Vannlokalitet_kode, year, water_type),
    hard001_yofj  |> distinct(Vannlokalitet_kode, year, water_type)
  ) |> distinct(Vannlokalitet_kode, year, .keep_all = TRUE), by = key) |>
  left_join(msmdi_thresholds, by = c("taxon_group", "water_type")) |>
  mutate(points = case_when(
    disappeared          ~ 0,
    is.na(depth_m)       ~ NA_real_,
    is.na(water_type)    ~ NA_real_,
    depth_m > p5         ~ 5,
    depth_m > p4         ~ 4,
    depth_m > p3         ~ 3,
    depth_m > p2         ~ 2,
    TRUE                 ~ 0
  ))

msmdi_recomputed <- msmdi_obs |>
  group_by(across(all_of(key))) |>
  summarise(
    route         = paste(sort(unique(route[route != "disappeared"])),
                          collapse = "; "),
    n_taxa_scored = sum(!is.na(points)),
    eqr_recomputed = if (sum(!is.na(points)) >= 3) {
      mean(points, na.rm = TRUE) / 5
    } else NA_real_,
    .groups = "drop"
  ) |>
  filter(!is.na(eqr_recomputed))
```

The comparison below is the evidence for §9.2. Our reimplementation can only look back to 2021 for previous occurrences, whereas the report series looks back to about 2009 [@lundsor2025]. A shorter lookback means fewer species scored zero and therefore a higher index, so we expect our values to sit slightly *above* the report — and that is what the bias column shows. The Vannmiljø values sit further above again, consistent with a still shorter or more incomplete occurrence record.

```{r tbl-msmdi-qa}
qa <- msmdi_recomputed |>
  inner_join(hard001 |> select(all_of(key), programme, eqr, eqr_vannmiljo),
             by = key) |>
  mutate(recomputed = round(eqr_recomputed, 2),
         d_used       = recomputed - eqr,
         d_vannmiljo  = recomputed - eqr_vannmiljo)

qa |>
  group_by(programme, route) |>
  summarise(
    n = n(),
    `within 0.05 of value used`  = sum(abs(d_used) <= 0.05, na.rm = TRUE),
    `bias vs value used`         = round(mean(d_used, na.rm = TRUE), 3),
    `bias vs Vannmiljø`          = round(mean(d_vannmiljo, na.rm = TRUE), 3),
    .groups = "drop"
  ) |>
  kable(caption = "Independent reimplementation of MSMDI against the two official series. For ØkoKyst the value used is the Skagerrak report; for Ytre Oslofjord the two are the same series.")
```

For the ØkoKyst stations the reimplementation sits close to the report series and well away from the Vannmiljø values. How much that choice matters:

```{r tbl-series-choice}
hard001 |>
  filter(programme == "OEKK", !is.na(eqr_vannmiljo)) |>
  summarise(
    `station-years` = n(),
    `mean difference` = round(mean(eqr - eqr_vannmiljo), 3),
    `largest difference` = round(max(abs(eqr - eqr_vannmiljo)), 2),
    `different condition class` = sum(as.character(neqr_class(eqr)) !=
                                        as.character(neqr_class(eqr_vannmiljo)))
  ) |>
  kable(caption = "Effect of using the Skagerrak report series rather than Vannmiljø for the ØkoKyst half of HARD_001")
```

### 9.5 Sea urchin grazing

The grazing correction cannot be computed (§5.1), so urchin abundance is reported as a caveat instead. Abundance is on the four-point transect scale, where 2 is *spredt* — the threshold at which the method considers grazing potentially significant.

```{r urchins}
urchins <- read_data("vm_hardbottom_fauna.csv") |>
  filter(Parameter_id %in% c("MHRFORSK", "MHDEKNSK"),
         VitenskapligNavn %in% c("Echinus esculentus",
                                 "Strongylocentrotus droebachiensis")) |>
  group_by(across(all_of(key))) |>
  summarise(max_urchin_abundance = max(value, na.rm = TRUE), .groups = "drop")

bind_rows(
  hard001 |> select(all_of(key), indicatorID),
  hard002 |> select(all_of(key), indicatorID)
) |>
  left_join(urchins, by = key) |>
  mutate(grazing_flag = case_when(
    is.na(max_urchin_abundance) ~ "no urchin record",
    max_urchin_abundance >= 2   ~ "urchins spredt or more - value is an upper bound",
    TRUE                        ~ "urchins below spredt"
  )) |>
  count(indicatorID, grazing_flag, name = "station_years") |>
  kable(caption = "Sea urchin abundance at the indicator stations. Where urchins are spredt or more, the indicator value should be read as an upper bound.")
```

## 10. Results

<!--#
Repeat the final results here. Typically this is a map or table of indicator values.
-->

```{r results-table}
results <- bind_rows(
  hard001 |> transmute(indicatorID, index, Vannlokalitet_kode, station_code,
                       station = Vannlokalitetsnavn, programme, ecoregion,
                       water_type, year, eqr, condition_class),
  hard002 |> filter(status == "computed") |>
    transmute(indicatorID, index, Vannlokalitet_kode, station_code,
              station = Vannlokalitetsnavn, programme, ecoregion,
              water_type, year, eqr, condition_class)
)
```

The headline figures, with the descriptive spread described in §5.2:

```{r tbl-results}
#| tbl-cap: "Indicator values by indicator and year. Spread is between monitoring stations, not uncertainty of a station value."
results |>
  group_by(indicatorID, index, year) |>
  summarise(stations = n(),
            median = round(median(eqr), 3),
            IQR = paste0(round(quantile(eqr, .25), 2), "-",
                         round(quantile(eqr, .75), 2)),
            range = paste0(round(min(eqr), 2), "-", round(max(eqr), 2)),
            .groups = "drop") |>
  kable()
```

```{r tbl-classes}
#| tbl-cap: "Number of stations per condition class."
results |>
  count(indicatorID, year, condition_class, .drop = FALSE) |>
  filter(n > 0 | condition_class %in% c("God", "Svært god")) |>
  pivot_wider(names_from = condition_class, values_from = n, values_fill = 0) |>
  kable()
```

HARD_001 has four years, so it can be shown as a trajectory. Each thin line is one station; the heavy line is the annual median. The horizontal band marks the good/moderate boundary at 0.6, which is the management-relevant threshold.

```{r fig-hard001}
#| fig-cap: "HARD_001 (MSMDI) by year. Thin lines are individual monitoring stations, the heavy line is the annual median across stations. Values below the dashed line fail the good/moderate boundary."
#| fig-width: 7
#| fig-height: 4.2
#| warning: false
median_001 <- hard001 |> group_by(year) |>
  summarise(eqr = median(eqr), .groups = "drop")

ggplot(hard001, aes(year, eqr)) +
  geom_hline(yintercept = 0.6, linetype = "dashed",
             colour = "grey45", linewidth = 0.4) +
  geom_line(aes(group = Vannlokalitet_kode),
            colour = "grey70", linewidth = 0.3) +
  geom_point(colour = "grey70", size = 1) +
  geom_line(data = median_001, linewidth = 1.1, colour = "#1E90D2") +
  geom_point(data = median_001, size = 2.4, colour = "#1E90D2") +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(x = NULL, y = "MSMDI (EQR)",
       title = "Macroalgae lower growth limit, Skagerrak",
       subtitle = "Annual median in blue; grey lines are individual stations") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank())
```

Condition class is what management acts on, so the class distribution is shown directly. Colours are the prescribed Norwegian class colours and are always paired with the class label.

```{r fig-classes}
#| fig-cap: "Number of monitoring stations per condition class. HARD_002 is available for 2024 only."
#| fig-width: 7.5
#| fig-height: 3.8
results |>
  mutate(panel = paste0(indicatorID, " - ", index)) |>
  count(panel, year, condition_class, .drop = FALSE) |>
  filter(!is.na(condition_class)) |>
  ggplot(aes(factor(year), n, fill = condition_class)) +
  geom_col(width = 0.72, colour = "white", linewidth = 0.6) +
  facet_wrap(~ panel, scales = "free_x") +
  scale_fill_manual(values = class_colours, name = "Condition class",
                    drop = FALSE) +
  labs(x = NULL, y = "Stations") +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor = element_blank(),
        legend.position = "bottom")
```

Finally, what a water-body series would look like for the part of the network where the mapping exists. This is the aggregation *Veileder 02:2018* actually prescribes, and it is shown here as a demonstration rather than as the indicator, because the mapping is missing for the Ytre Oslofjord stations.

```{r tbl-waterbody}
#| tbl-cap: "Water-body means for 2024, for stations where a water body mapping exists. Not the indicator value; see section 8."
results |>
  filter(year == 2024) |>
  left_join(station_meta |> select(Vannlokalitet_kode, water_body), by = "Vannlokalitet_kode") |>
  filter(!is.na(water_body)) |>
  group_by(indicatorID, water_body) |>
  summarise(stations = n(), neqr = round(mean(eqr), 3), .groups = "drop") |>
  mutate(condition_class = neqr_class(neqr)) |>
  arrange(indicatorID, neqr) |>
  kable()
```

## 11. Export file

<!--#
Optional: Display the code (don't execute it) or the workflow for exporting the indicator values to file.
-->

The station-level indicator values, with the flags that qualify them, would be exported as follows. Coordinates are UTM33 (EPSG:25833) as supplied by Vannmiljø.

```{r export}
#| eval: false
export <- results |>
  left_join(stations |> select(Vannlokalitet_kode, utm33_x, utm33_y),
            by = "Vannlokalitet_kode") |>
  left_join(station_meta |> select(Vannlokalitet_kode, water_body),
            by = "Vannlokalitet_kode") |>
  left_join(hard002 |> select(Vannlokalitet_kode, year, flag_old_point_scale,
                              flag_partial_sjosone),
            by = c("Vannlokalitet_kode", "year"))

readr::write_csv(export, "../data/hard_001_002_indicator_values.csv")

# As a georeferenced point layer
sf::st_as_sf(export, coords = c("utm33_x", "utm33_y"), crs = 25833) |>
  sf::st_write("../data/hard_001_002_indicator_values.gpkg", delete_dsn = TRUE)
```

## 12. Author contributions

<!-- Fill in CReDIT and MeRIT statments to clarify the contributions from individual authors. -->

**CRediT statement**

* Conceptualization: DØH
* Methodology: DØH
* Data curation: DØH
* Formal analysis: DØH
* Software: DØH
* Validation: DØH
* Visualization: DØH
* Writing – original draft: DØH

**MeRIT attribution**

**Study design**
The two indices were designed by NIVA for the Norwegian Environment Agency and are documented in Veileder 02:2018 and Fagerli et al. (2025). This submission documents their use as ecosystem condition indicators; it does not introduce new index methodology.

**Data collection / fieldwork**
Field data were collected by NIVA and partners under the ØkoKyst programme, and by NIVA for Fagrådet for Ytre Oslofjord, between 2021 and 2024.

**Data processing**
The Vannmiljø extract was filtered and the report tables transcribed by DØH.

**Statistical analysis**
Index assembly, the independent reimplementation of MSMDI used as a quality check, and the comparison between the two official MSMDI series were done by DØH.

**Reproducibility**
Not yet independently reproduced. The code review is pending.


::: {.callout-tip collapse="true"}
## Session Info
<!-- You can leave this last part as it is, unless you have a good reason to alter it. -->

This information is mainly aimed at developers that need to recreate the R environment that produced the html.
Important note: this information is about the R session that was running when the html was rendered, and if some of the code chunks were not evaluated when the html was rendered, they may have been executed in a different environment.

```{r sessionInfo}
#| echo: false
sessionInfo()
```

:::
